---
title: "I h8 naming things"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## load packages

loading pakages required for study

```{r}
library(readxl)   
library(tidyverse)
```

## read the finches data

Have R associate the excel file with the name "finches"

```{r}
finches <- read_excel("finches_data.xlsx")
```

## print the finches tibble in the console

make the data readable in the console

```{r}
finches
```

## take a quick look at all the variables in the dataset

shows all the variables laid out

```{r}
glimpse(finches)
```

## histogram ---------------------------------------------------------------

# histogram of beak length, grouped by survival, with labels

creates a histogram so the data is more visually appealing and easier to understand.

```{r}
ggplot(
  data = finches,                     
  mapping = aes(x = beak_length,      
                fill = outcome)       
) +
  geom_histogram(bins = 14) +         
  facet_wrap(~ outcome, ncol = 1) +   
  guides(fill = FALSE) +              
  labs(
    title = "Figure 1.",              
    x = "Beak Length (mm)",          
    y = "Number of Birds")   
```

## summarize ---------------------------------------------------------------

# summarize the dataset by outcome (survived vs. died)

Creates a simple summary of the data.
```{r}
beak_length_grouped_summary <- 
  finches %>% 
  group_by(outcome) %>% 
  summarize(mean = mean(beak_length),
            sd = sd(beak_length),
            n = n()) %>% 
  mutate(sem = sd / sqrt(n),
         upper = mean + 1.96 * sem,
         lower = mean - 1.96 * sem)
```

## print the results in the console

show the results in a comprehesible way
```{r}
beak_length_grouped_summary
```

## bar chart ---------------------------------------------------------------

creates a bar chart of mean beak lengths.

```{r}
ggplot(
  data = beak_length_grouped_summary,   
  mapping = aes(x = outcome,            
                y = mean,               
                fill = outcome)         
) +
  geom_col() +                         
  geom_errorbar(                        
    mapping = aes(ymin = lower,         
                  ymax = upper),        
    width = .3                          
  ) +
  guides(fill = FALSE) +                
  labs(
    title = "Figure 2.",                
    x = "Survival Outcome",             
    y = "Beak Length (mm)")
```


## t-test 

created objects based on the 2 outcomes: died and survived.  first is died.

```{r}
beak_length_died <-
  finches %>%
  filter(outcome == "died") %>%  
  pull(beak_length)               
```

## print the new object in the console

show the raw data of "died".

```{r}
beak_length_died
```

## create object

create objec for beak lengths for birds that survived

```{r}
beak_length_survived <-
  finches %>% 
  filter(outcome == "survived") %>% 
  pull(beak_length)
```

## print the results in the console

show raw data of "survived".
```{r}
beak_length_survived
```

## perform a two-sample t-test assuming unequal variances

do a T-test using the 2 objects.
```{r}
t.test(beak_length_died, beak_length_survived)
```
